
\chapter{Porting the QPF formalization from Lean3 to Lean4}%
\label{ch:porting}

Lean is primarily developed by Leonardo de Moura \etal{} at Microsoft Research.~\cite{demouraLeanTheoremProver2015, avigadTheoremProvingLean} As an in-progress research project, it explicitly provides no stability guarantees, and as a result, Lean 4 is backwards incompatible with Lean 3.
This means that the QPF formalizations by Avigad et al. \etal{}
(\cite{avigadDataTypesQuotients2019a}) were not directly usable. The current chapter details my efforts to port these formalizations to Lean 4.



Lean 3 has a comprehensive, community-maintained mathematics library, Mathlib (\cite{themathlibcommunityLeanMathematicalLibrary2020}), 
which also functions as its unofficial, de-facto standard library.
Mathlib4 is the port of this library to Lean 4 (\cite{mathlib4}),
but large parts were still missing at the time I started my thesis project.
In particular, Avigad \etal{} contributed their formalizations to Mathlib, but these were far from
being ported.



\subsection*{Upstreaming}
It should be noted that at the time, the maintainers of mathlib4 were not yet interested in 
community-ported files. This has since changed and I am currently also in the process of
contributing my porting work to Mathlib4, making it more widely available.




\section{Mathport and Changes from Lean 3}
Lean 4 is not just a new version of the language, most of the code base has been rewritten.

There were some changes in the syntax and naming convention, but most importantly, 
the metaprogramming in Lean 4 has been completely reworked.

To help with porting code, the mathlib community developed the mathport tool (\cite{mathport}), which 
(best-effort) translates Lean 3 source code into Lean 4 source code. This takes care of superficial 
syntax and naming changes, but the translation is far from perfect.
In particular, mathport did reasonably well in translating declaration signatures, 
but proofs did not translate well.

This is primarily because mathport intentionally does not deal with metaprogramming code, and a lot
of tactics (proof automation meta-programs) provided by Mathlib were still missing from Mathlib4.


\section{Implicit arguments and other inference}
Lean relies heavily on automated inference to reduce verbosity, falling back on explicitly provided
values or annotations when inference gets things wrong.
Lean 4 will sometimes infer something different, and cause an error.

By default, the pretty printer used to show intermediate proof states and types on hover in vscode will
omit any implicit arguments in the term, to closely resemble whatever was written in the source code.
However, we can instruct it to be more verbose:
\begin{leancode}
  set_option pp.implicit true -- print implicits, or
  set_option pp.all true      -- set all options ()
\end{leancode}
These work in both Lean 3 and Lean 4, so it makes it easy to do a side-by-side comparison.

For example, in the following theorem Lean 4 couldn't figure out the right types for \lean{f ::: g} and 
\lean{f' ::: g'}, so explicit type annotations, which look like \lean{(f ::: g : \_)}, had to be added.
\begin{leancode}
  theorem append_fun_inj {α α' : Typevec n} {β β' : Type _} 
                         {f f' : α ⟹ α'} {g g' : β → β'} :
  (f ::: g : (α ::: β) ⟹ _) = (f' ::: g' : (α ::: β) ⟹ _) → f = f' ∧ g = g'
\end{leancode}

Similarly, Lean does inference sligthly different for so called \emph{eliminators}.
Lean 3 (and more recent versions of Lean 4) have a \leanm{@[elab_as_eliminator]} attribute on user-defined
functions that instructs Lean to use the specialized eliminator inference, but the version we're using
does not have that.

This came up, e.g., in the proof of \lean{Cofix.bisim\_aux}. The Lean 3 proof simply used
\begin{leancode}    
  apply quot.inductionOn c
\end{leancode}  
Where \lean{Quot.inductionOn} is marked with \lean{elab\_as\_eliminator}.
In Lean 4 we had to explicitly provide the \emph{motive} (which determines the induction hypothesis), 
since it was being inferred incorrectly.
\begin{leancode}
  apply Quot.inductionOn (motive := fun c => ∀b, r c b → 
                                      Quot.lift (Quot.mk r') h₁ c 
                                      = Quot.lift (Quot.mk r') h₁ b) 
            c
\end{leancode}

The tricky part is that the proof generally doesn't throw an error at \lean{Quot.inductionOn} when it
infers the wrong motive; it just produces a different proof state, which trips up later tactics.
Catching this involved stepping through both the Lean 3 and Lean 4 to find out where, exactly, the proof states
diverged, and then explicitly providing the desired motive.

This problem was thus hard to diagnose, but relatively easy to remedy.


\section{Missing Tactics} 
Conversely, the issue of missing tactics was easy to diagnose, but harder to fix.
The long-term best solution would have been to implement all missing tactics.
However, there were quite a few, and implementing them would probably have taken a long time.

Instead, proofs that used such missing tactics were changed to use the lower-level tactics that were
available, making these proofs a bit longer and more verbose.

Notably, Mathlib4 has made much progress in implementing more tactics since, 
so presumably I can change most of these proofs back to use the higher-level tactics when
making the changes needed to contribute them back to Mathlib4.



